An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-03-26 EconGrowth-R-Functions.Rnw
Changed a few limits to the CI vs. parameter graphs.
Added a new function in the CD functions section to make a partial residuals plot: residuals vs. e for the Cobb-Douglas without energy model.

####################
# Returns an nls Cobb-Douglas model for the country specified
# energyType is one of "Q", "X", or "U".
##
cdeModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We need to do the Cobb-Douglas fit with the desired energy data.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function. 
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish guess values for lambda, alpha, and beta.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.9
  betaGuess <- 1.0 - alphaGuess
  aGuess <- alphaGuess
  bGuess <- alphaGuess + betaGuess
  start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
  # Now actually do the fit, using the column name "iEToFit".
  # gamma is a free parameter.
  # Reparameterize to ensure that we meet the constraints:
  # * alpha + beta + gamma = 1.0.
  # * alpha, beta, and gamma are all between 0.0 and 1.0.
  # To do this, we reparameterize as
  # * 0 < a < 1
  # * 0 < b < 1
  # * alpha = min(a, b)
  # * beta = b - a
  # * gamma = 1 - max(a, b)
  model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
  modelCDe <- nls(formula=model, data = dataTable, start = start,
                  control = nls.control(maxiter = 200, 
                                        tol = 1e-05, 
                                        minFactor = 1/1024, 
                                        printEval=FALSE, #Tells whether to print details of curve fit process.
                                        warnOnly=TRUE),
                  #Include the next 3 lines to fit with constraints.
                  #algorithm = "port",
                  #lower = list(lambda=-Inf, a=0, b=0),
                  #upper = list(lambda= Inf, a=1, b=1)
                  )
  return(modelCDe)
}

@@

@@

@@

@@

@@

@@

@@

@@

                             tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.

>>>>

####################
# Returns an nls Cobb-Douglas model for the country specified
# energyType is one of "Q", "X", or "U".
##
cdeModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  # We need to do the Cobb-Douglas fit with the desired energy data.
  # To achieve the correct fit, we'll change the name of the desired column
  # to "iEToFit" and use "iEToFit" in the nls function. 
  dataTable <- replaceColName(dataTable, energyType, "iEToFit")
  # Establish guess values for lambda, alpha, and beta.
  lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
  alphaGuess <- 0.9
  betaGuess <- 1.0 - alphaGuess
  aGuess <- alphaGuess
  bGuess <- alphaGuess + betaGuess
  start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
  # Now actually do the fit, using the column name "iEToFit".
  # gamma is a free parameter.
  # Reparameterize to ensure that we meet the constraints:
  # * alpha + beta + gamma = 1.0.
  # * alpha, beta, and gamma are all between 0.0 and 1.0.
  # To do this, we reparameterize as
  # * 0 < a < 1
  # * 0 < b < 1
  # * alpha = min(a, b)
  # * beta = b - a
  # * gamma = 1 - max(a, b)
  model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
  modelCDe <- nls(formula=model, data = dataTable, start = start,
                  control = nls.control(maxiter = 200, 
                                        tol = 1e-05, 
                                        minFactor = 1/1024, 
                                        printEval=FALSE, #Tells whether to print details of curve fit process.
                                        warnOnly=TRUE),
                  #Include the next 3 lines to fit with constraints.
#                   algorithm = "port",
#                   lower = list(lambda=-Inf, a=0, b=0),
#                   upper = list(lambda= Inf, a=1, b=1)
                  )
  return(modelCDe)
}

@@

#################
# Creates a plot with y residuals vs. energy (e). The y residuals are from a Cobb-Douglas model without energy.
# The energy term on the x axis is whatever type is specified in the calling funciton
partialResidualPlot <- function(countryAbbrev, energyType, textScaling=1.0){
  # Get the model
  cdModel <- cdModel(countryAbbrev)
  # Get the residuals
  cdResid <- resid(cdModel)
  cdResid <- data.frame(cdResid)
  cdResid <- padRows(countryAbbrev=countryAbbrev, df=cdResid)
  # Load data
  data <- loadData(countryAbbrev)
  data <- cbind(data, cdResid)
  data <- replaceColName(dataTable=data, factor=energyType, newName="iEToFit")
  plot <- xyplot(cdResid ~ iEToFit, data=data,
                 type=c('p','smooth'),
                 scales=list(cex=scaleTextSize*textScaling, #controls text size on scales.
                             tck=scaleTickSize), #controls tick mark length. < 0 for inside the graph.
                 xlab=list(label=paste("$", tolower(energyType), "$", sep=""), cex=textScaling),
                 ylab=list(label="$y$ residuals", cex=textScaling)
                 )
  return(plot)  
}
@

@@

  # Also, set limits on the x and y axes if needed

@@

    xLimits <- c(0.0, 0.1); atX=c(0.0, 0.05, 0.1)
    yLimits <- c(0.0, 0.2); atY-c(0.0, 0.1, 0.2)

@@

    xLimits <- c(-0.1, 1.5); atX <- c(0.0, 0.5, 1.0, 1.5)
    yLimits <- c(0.0, 1.5); atY <- c(0.0, 0.5, 1.0, 1.5)

@@

    xLimits <- c(-0.1, 1.1); atX <- c(0.0, 0.5, 1.0)
    yLimits <- c(0.0, 1.5); atY <- c(0.0, 0.5, 1.0, 1.5)

@@

    xLimits <- c(-0.1, 1.1); atX <- c(0.0, 0.5, 1.0)
    yLimits <- c(-0.1, 1.5); atY <- c(0.0, 0.5, 1.0, 1.5)

@@

    xLimits <- c(-0.1, 1.1); atX <- c(0.0, 0.5, 1.0)
    yLimits <- c(-0.1, 2.0); atY <- c(0.0, 1.0, 2.0)

@@

                             x=list(at=atX),
                             y=list(at=atY)
                 xlim=xLimits, #x axis limits

### 2013-03-26 EconGrowthSeminar.Rnw
Added a new function in the CD functions section to make a partial residuals plot: residuals vs. e for the Cobb-Douglas without energy model.

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?

    \pause

    No known studies of comparing economic growth models for various types of economies on a rigorous statistical basis.
  \end{center}
\end{frame}

@@

\begin{frame}{Coordinates of Analysis} %----------------------------
  \begin{itemize}
    \item{Country/Economy: 9 \\US, UK, JP, CN, ZA, SA, IR, TZ, ZM}
    \item{Economic Growth Model: 4 \\SF, CD, CES, LINEX}
    \item{Type of Energy: 3 \\Q, X, U}
  \end{itemize}
  Simplification: 1 type of energy (Q)
\end{frame}

@@

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits historical data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint: $\alpha + \beta + \gamma = 1$}
    \item{Respect absolute constraints: $0 \leq \alpha, \beta, \gamma \leq 1$}
  \end{itemize}
\end{frame}

@@

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with Energy} %----------------------------
<<CI_gamma_vs_gamma_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="Q")
@
\end{frame}

@@

\begin{frame}{95\% Confidence Intervals}{CES with $q$} %----------------------------
<<CESq_Params_Graph, fig.width=presentationGraph1ColWidth*1.8, fig.height=presentationGraph1RowHeight*1.8, eval=TRUE>>=
createCESParamsGraph("Q")
@  
\end{frame}

@@

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES} %----------------------------
<<CI_delta_vs_delta_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CES", param="delta", energyType="Q")
@
\end{frame}

@@

\begin{frame}{$CI_{\sigma}$ vs. $\sigma$ for CES} %----------------------------
<<CI_sigma_vs_sigma_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CES", param="sigma", energyType="Q")
@
\end{frame}

@@

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Estimate statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \vspace{\baselineskip}
    \item{Complete paper}
  \end{itemize}  
\end{frame}

>>>>

\begin{frame}{The Question} %----------------------------
  \begin{center}
    Which economic growth model best describes (statistically) the role of energy in economic growth for different types of economies?
    \vspace{\baselineskip}
    \pause

    No known comparative studies of economic growth models for various types of economies on a rigorous statistical basis.
  \end{center}
\end{frame}

@@

\begin{frame}{Coordinates of Analysis} %----------------------------
  \begin{itemize}
    \item{Country/Economy: 9 \\US, UK, JP, CN, ZA, SA, IR, TZ, ZM}
    \item{Economic Growth Model: 4 \\SF, CD, CES, LINEX}
    \item{Type of Energy: 3 \\Q, X, U}
  \end{itemize}
  \vspace{\baselineskip}
  \pause

  \begin{center}Simplification: 1 type of energy ($Q$) for this presentation\end{center}
\end{frame}

@@

\begin{frame}{Parameter Estimation for Cobb-Douglas With Energy} %----------------------------
  \begin{center}\cdeEqn\end{center}
  \begin{itemize}
    \item{Objective: determine $\lambda$, $\alpha$, $\beta$, and $\gamma$ that best fits historical data}
    \item{Almost identical to Cobb-Douglas without energy}
    \item{Perform grid search over $\gamma$}
    \item{Re-estimate $\lambda$, $\alpha$, and $\beta$ at each value of $\gamma$}
    \item{Respect sum constraint: $\alpha + \beta + \gamma = 1$}
    \item{Respect absolute constraints: $0 \leq \alpha, \beta, \gamma \leq 1$} (not yet)
  \end{itemize}
\end{frame}

@@

\begin{frame}{$CI_{\gamma}$ vs. $\gamma$ for Cobb-Douglas with $Q$} %----------------------------
<<CI_gamma_vs_gamma_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CDe", param="gamma", energyType="Q")
@
\end{frame}

@@

\begin{frame}{95\% Confidence Intervals}{CES with $q$} %----------------------------
<<CESq_Params_Graph, fig.width=presentationGraph1ColWidth*1.8, fig.height=presentationGraph1RowHeight*1.8>>=
createCESParamsGraph("Q")
@  
\end{frame}

@@

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES with $Q$} %----------------------------
<<CI_delta_vs_delta_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CES", param="delta", energyType="Q")
@
\end{frame}

@@

\begin{frame}{$CI_{\sigma}$ vs. $\sigma$ for CES with $Q$} %----------------------------
<<CI_sigma_vs_sigma_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5, eval=TRUE>>=
CIvsParamPlot(model="CES", param="sigma", energyType="Q")
@
\end{frame}

@@

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Estimate statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \item{Incorporate all constraints for Cobb-Douglas model}
    \vspace{\baselineskip}
    \item{Complete paper}
  \end{itemize}  
\end{frame}
